---
title: "intro_to_COPDs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro_to_COPDs}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(copdr)
```

# 1. How are probability distribution functions used to analyze earthquakes?

In general, when scientists are trying to determine the size of past
earthquakes, they need to know how much slip occurred in a single
earthquake. This is because the magnitude of total offset is correlated
to the moment magnitude/energy released during an earthquake (e.g., the
Richter Scale). To determine the amount of total offset that occurs
during earthquakes, we must find linear features that intersect the
surface trace of a fault, and reconstruct the feature by 'backslipping'
or undoing the motion that deformed that originally-linear feature.
There are fantastic historic examples of roads and fences being split
apart by earthquakes, but we may also use naturally sub-linear geologic
features such as glacier moraines, streams, and ridges.

## 1.1 Example offset from the field
Southeastern California has many examples of well-exposed linear features, or 'piercing points' that we can reconstruct.
In Figure 1, a single point bar is right-laterally offset by a single fault trace.

```{r fig-offset-bar, echo=FALSE, out.width = "75%", fig.cap = "Figure 1: A right-laterally offset bar in an alluvial fan. Dashed white lines show the bar edges, the red line indicates the fault trace, and yellow arrows show the sense of slip. Panamint Valley, California. Modified from LaPlante (2022)."}
fig1url = "https://github.com/AubreyLaPlante/external_figs/blob/main/RLoffset.png?raw=true"
knitr::include_graphics(fig1url)
```
***
Assuming that this was one feature originally, we can measure how much
one side of the bar moved relative to the other by measuring the
distance between the southern edge of the western bar segment and the
southern edge of the eastern bar segment. We can also measure the
distance between the northern edge of the western bar segment and the
northern edge of the eastern bar segment. There is also the crest of the
bar, representing its maximum height at a central E-W axis, that
provides another lateral offset measurement. As you might imagine, there
is considerable uncertainty within where these bar edges and crest lines
are, and this can change the measured lateral offset of this feature,
depending on scientific interpretations. Confused yet?

## 1.2 Graphical example of offset reconstruction
Figure 2 is a graphical representation of a right-laterally offset bar,
similar to the geomorphic feature shown in Figure 1. 

```{r fig-reconstruct-bar-schematic, echo=FALSE, out.width = "100%", fig.cap = "Figure 2: A schematic diagram showing the reconstruction, or 'backslipping', of a point bar. (a) The fault (solid line) and the geomorphic profile of the bar on the western side of the fault (A-A’) and the eastern side of the fault (B-B’). (b) Topographic profiles A-A’ and B-B’ before lateral and vertical shift. (c) Topographic profiles A-A’ and B-B’ after lateral and vertical shift. (d) By mathematically restoring the offset geomorphic marker, a Matlab code generates a lateral shift-correlation plot that predicts a most likely displacement for a piercing point. The vertical offset is found by finding the amount of vertical shift necessary to achieve a correlation coefficient of 1 on the lateral profile shift line. Figure from LaPlante (2022)."}
fig2url= "https://github.com/AubreyLaPlante/external_figs/blob/main/LaPlante2022_PDFs.png?raw=true"
knitr::include_graphics(fig2url)
```
What Figure 2d shows is that when we estimate the amount of slip on a fault from an
offset feature by 'backslipping' that feature, we get a most-likely
value (where correlation = 1), bounded by a maximum and a minimum slip
value (where correlation = 0). This curve approximates a normal/Gaussian probability distribution function (PDF) where the width of this distribution is related to the uncertainty of the offset measurement. We can quantify that uncertainty in a number
of ways including the standard deviation, the coefficient of variation,
or confidence scaling. Confidence scaling can vary across different
scientific studies, and is used to more-strongly weight the more
sharply-defined piercing points (linear features) when we combine/sum
multiple PDFs together to create a cumulative-offset probability distribution function.

***

# 2.  What are Cumulative-offset Probability Distribution (COPD) functions?

First of all, what are cumulative-offset probability distribution (COPD)
functions, and how are they used in earthquake offset analyses? COPDs are the summed total of all the normal PDFs, across the entire length of a fault zone. One commonly used example is the well-defined COPD
plots from the Carrizo Plain segment of the San Andreas fault in
California, US. Similar to the example shown above in Figure 2, Zielke et al.,
(2010) illustrates one example of right-lateral offset of a single
drainage. In this example, the PDF peak occurs at around 5 m, with an
uncertainty of ± \~0.6 - 0.7 m.


```{r fig-zielke-channel, echo=FALSE, out.width = "100%", fig.cap = "Figure 3: A single offset displacement marker from the Carrizo segment of the San Andreas fault, in southern California. (a) The aerial view of the offset channel from a digital elevation model. (b) Red and blue cross sections of the offset channel, parallel to the fault plane. (c) The resulting right-lateral offset PDF for this channel. Figure from Zielke et al., (2010)."}
fig3url = "https://github.com/AubreyLaPlante/external_figs/blob/main/Zielke2010_channels.png?raw=true"
knitr::include_graphics(fig3url)
```

***
Zielke et al., (2010) then combines dozens of PDFs along this segment of the
San Andreas fault, as seen in Figure 4. 


```{r fig-zielke-copd, echo=FALSE, out.width = "100%", fig.cap = "Figure 4: Individual PDFs from a ~50 km extent of the Carrizo fault segment of the southern San Andreas fault. All PDFs are summed together to create a COPD plot, which is useful in interpreting the magnitude of total slip in past earthquakes. (a) Individual PDFs are plotted in space, with distance from north to south on the X axis and total offset magnitude on the Y axis. PDFs are colored by average slip magnitude, with orange representing the clustering of the least amount of slip and light blue representing the clustering of the largest magnitude of slip. (b) All PDFs are summed together to isolate clusters of total offset magnitudes. Theoretically, each peak represents the displacement associated with a different number of earthquakes, assuming that each earthquake produced the same magnitude of slip. Therefore, orange curves represent one earthquake, purple curves represent two earthquakes, yellow represents three earthquakes, etc. Figure from Zielke et al., (2010)."}
fig4url = "https://github.com/AubreyLaPlante/external_figs/blob/main/Zielke2010_carrizo.png?raw=true"
knitr::include_graphics(fig4url)
```
***
This combined/summed PDF plot (Figure 4b) is the cumulative-offset probability distribution (COPD) function. Each of
these individual PDFs have different local offset maxima because each
offset marker records a different number of earthquakes. This is
consistent with what we would expect, such that the oldest geomorphic
features have been displaced by many earthquakes, while the youngest
generations of Holocene alluvium may have only experienced the most
recent earthquake. Assuming that earthquakes on the southern San Andreas
fault are perfectly elastic, such that this fault repeatedly produces the same size earthquake on the same size rupture patch in subsequent events, we would also expect to see regular PDF peak spacing from subsequent earthquakes producing the same total
magnitudes of slip.

In Figure 4b, we see peaks in the COPD plot of 5m, 10m, 15m, and so on.
This is useful, and consistent with our assumptions. It appears that
each earthquake on the Carrizo segment produces about \~5m of
right-lateral strike slip motion. Using this number for average
displacement we can use scaling relationships (Wells and Coppersmith,
1994) to determine that this amount of average displacement corresponds
to a rupture length of \~136 km and an earthquake magnitude of Mw \~7.7.
COPDs can be really useful to estimate future seismic hazards!
